---
title: "Capstone Project"
author: "Cosmel Vilallboos"
date: '2022-06-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up the evironment


```{r loading pacakges, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)

```

##  STEP 1: COLLECT DATA
### Loading the data
```{r warning=FALSE}

m1_2021 <- read.csv("202101-divvy-tripdata.csv")
m2_2021 <- read.csv("202102-divvy-tripdata.csv")
m3_2021 <- read.csv("202103-divvy-tripdata.csv")
m4_2021 <- read.csv("202104-divvy-tripdata.csv")
m5_2021 <- read.csv("202105-divvy-tripdata.csv")
m6_2021 <- read.csv("202106-divvy-tripdata.csv")
m7_2021 <- read.csv("202107-divvy-tripdata.csv")
m8_2021 <- read.csv("202108-divvy-tripdata.csv")
m9_2021 <- read.csv("202109-divvy-tripdata.csv")
m10_2021 <- read.csv("202110-divvy-tripdata.csv")
m11_2021 <- read.csv("202111-divvy-tripdata.csv")
m12_2021 <- read.csv("202112-divvy-tripdata.csv")

```

## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
### Stack individual quarter's data frames into one big data frame
```{r warning=FALSE}
all_trips <- bind_rows(m1_2021,m2_2021,m3_2021,m4_2021,m5_2021,m6_2021,m7_2021,m8_2021,m9_2021,m10_2021,m11_2021,m12_2021)
```

### Remove lat, long fields as this data was dropped beginning in 2020
```{r warning=FALSE}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```


## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
### Displays the statistical summary of the whole data frame
```{r warning=FALSE}
summary(all_trips)
```

### Checking the amount of users per label
```{r warning=FALSE}
table(all_trips$member_casual)
```

### Add columns that list the date, month, day, and year of each ride
```{r warning=FALSE}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

### Add a "ride_length" calculation to all_trips (in minutes) and apply round to get back 2 digits
```{r warning=FALSE}
all_trips$ride_length <- round(difftime(all_trips$ended_at,all_trips$started_at,units = "mins"),2)
```


### Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r warning=FALSE}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

### Remove "bad" data
```{r warning=FALSE}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```


### Assignin the data that we need

```{r warning=FALSE}
data("all_trips_v2")
```

# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
### Summary(all_trips_v2$ride_length)
```{r warning=FALSE}
summary(all_trips_v2$ride_length)
```

### Compare members and casual users
```{r warning=FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

### Compare bykes per type
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$rideable_type, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$rideable_type, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$rideable_type, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$rideable_type, FUN = min)

```

### Re arrange the days to show in order 
```{r warning=FALSE}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### Visualization including rideable_types
```{r warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday,rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday,rideable_type)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = rideable_type)) +
  geom_col(position = "dodge")
```

### Visualization for average duration
```{r warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Visualize the number of rides by rider type
```{r warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
